---
title: "STAT-632_Project"
author: "Sumasree Simhadri"
date: "2023-04-14"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#installing caret package to use for splitting the dataset
#install.packages("caret")
```

```{r}
library(MASS)
library(car)
```

```{r}
#load the data set
whodata<-read.csv("Life Expectancy Data.csv")
#removing t=rows from the dataset that have NA value
whodata<-na.omit(whodata)
head(whodata)
```

```{r}
#building a multi-linear regression model with life expectancy as the response variable 
model1<-lm(Life.expectancy~Year+Alcohol+Total.expenditure+GDP+Income.composition.of.resources+Schooling+Population+percentage.expenditure+HIV.AIDS+Hepatitis.B+Measles,data = whodata)
summary(model1)

```

$\hat{Life.expectancy}<-\hat{\beta_0}+\hat{\beta_1}\hat{Year}+\hat{\beta_2}\hat{Alcohol}+\hat{\beta_3}\hat{Total.expenditure}+\hat{\beta_4}\hat{GDP}+\hat{\beta_5}\hat{Income.composition.of.resources}+\hat{\beta_6}\hat{Schooling}+\hat{\beta_7}\hat{Population}+\hat{\beta_8}\hat{Percentage.expenditure}+\hat{\beta_9}\hat{HIV.AIDS}+\hat{\beta_{10}}\hat{Hepatitis.B}+\hat{\beta_{11}}\hat{Measles}$

```{r}
#checking the 95% confodent interval for the coefficients of the variables
confint(model1)
```

We are 95% confident that the coefficient of GDP,Population,Measles can be zero,indicating that they are not related to the response variable.The p values of these 3 variables are greater than $\alpha$.

```{r}
library(ggplot2)
df <- data.frame(whodata)
#plot( whodata$percentage.expenditure,whodata$Life.expectancy, colo)
ggplot( df,aes(x = whodata$Schooling, y = Life.expectancy, )) + 
  geom_point() +
  labs(x = "Schooling", y = "life expectancy", title = "Relationship between Life expectancy and Schooling") +
  theme_bw()  
```

**Performing F-test to see if the coefficients are zero or not**

$H_0:\beta_4=\beta_7=\beta_{11}=0$

$H_1:\beta_4=\beta_7=\beta_{11}\neq0$

```{r}
#building an other model by removing the 3 variables-GDP,Population,Measles
model2<-lm(Life.expectancy~Year+Alcohol+Total.expenditure+Income.composition.of.resources+Schooling+percentage.expenditure+HIV.AIDS+Hepatitis.B,data = whodata)

model1<-lm(Life.expectancy~Year+infant.deaths+Alcohol+Total.expenditure+GDP+Income.composition.of.resources+Schooling+Population+percentage.expenditure+HIV.AIDS+Hepatitis.B+Measles,data = whodata)
#performing F test to compare the models 
anova(model2,model1)
```

Since the pvalue of f-test \>0.05,we fail to reject the null hypothesis. So we can remove those 3 predictors, Therefore,GDP,Population,Measles are insignificant .

```{r}
summary(model2)
```

Checking the l assumptions-

```{r}
#linearlity
pairs(Life.expectancy~Year+Alcohol+Total.expenditure+Income.composition.of.resources+Schooling+percentage.expenditure+HIV.AIDS+Hepatitis.B,data = whodata)
```

As seen,the percentage.expenditure is not linearly related to the response variable.Transformation is applied

```{r}
#removing the rows that have 0 as the percentage expenditure.
#This new data is stored in whonew
whonew<-whodata[whodata$percentage.expenditure!=0,]
```

```{r}
#building a new model after transforming the predictor variable.
model3<-lm(Life.expectancy~Year+Alcohol+Total.expenditure+Income.composition.of.resources+Schooling+log(percentage.expenditure)+HIV.AIDS+Hepatitis.B,data = whonew)
summary(model3)
```

Checking the assumptions for the new model

```{r}
#linearity
pairs(Life.expectancy~Year+Alcohol+Total.expenditure+Income.composition.of.resources+Schooling+log(percentage.expenditure)+HIV.AIDS+Hepatitis.B,data = whonew)
```

The linearity condition is satisfied.

```{r}
#normality
qqnorm(rstandard(model3))
qqline(rstandard(model3))
```

```{r}
#variability
plot(predict(model3), rstandard(model3),xlab="Fitted Values", ylab="Standardized Residuals")
abline(h=0)
```

The points in the residual plot look randomly scattered and show no obvious patterns, indicating that the\
assumptions are reasonably satisfied(linearity and constant variance).From the QQ plot we can say that the normality has been satisfied too since the qqline passes through the majority of the points.

```{r}
p <- 8#number of predictor variables in the model3
n <- nrow(whodata)#number of observations/rows
#scatter plot of h-values and standardized residuals
plot(hatvalues(model3), rstandard(model3),xlab='Leverage', ylab='Standardized Residuals',main = 'Outliers and Leverage points ')

abline(h=c(-2,2),v = 2*(p+1)/n, lty=4)


```

```{r}
#detecting multi colinearity using vif
round(vif(model3),2)
```

The variation inflation factors of the predictor variables are less than 5.

```{r}
#performing the step() function to select a model by using AIC
stepmodel<-stepAIC(model1,direction = "both")
summary(stepmodel)
```

The model suggested by the step() function using AIC is slightly different from the model3.

Training and testing is performed on both these models.

```{r}
library(caret)
```

```{r}
#setting a seed
set.seed(100)
#using the createDataPartition() function to split the dataset into training and testing data (70:30)
index<-createDataPartition(y=whonew$Life.expectancy,p=0.7,list=FALSE)
train<-whonew[index,]
test<-whonew[-index,]
```

```{r}
#model3 built by the training data
finalmodel<-lm(Life.expectancy~Year+Alcohol+Total.expenditure+Income.composition.of.resources+Schooling+log(percentage.expenditure)+HIV.AIDS+Hepatitis.B,data = train)
summary(finalmodel)
```

```{r}
#suggested model built on the train data
suggestedmodel<-lm(formula = Life.expectancy ~ Year + infant.deaths + Alcohol + 
    Total.expenditure + Income.composition.of.resources + Schooling + 
    percentage.expenditure + HIV.AIDS + Hepatitis.B, data = train)
summary(suggestedmodel)
```

```{r}
#using the predict function  on the finalmodel to predict the life.expectancy given the test data
predict_Life.expecentancy_final<-predict(finalmodel,test)
test["predicted.by.finalmodel"]<-predict_Life.expecentancy_final
```

```{r}
plot(test$predicted.by.finalmodel,test$Life.expectancy)
abline(lm(test$predicted.by.finalmodel~test$Life.expectancy))
```

```{r}
#using the predict function  on the suggestedmodel to predict the life.expectancy given the test data
predict_Life.expecentancy_suggested<-predict(suggestedmodel,test)
test["predicted.by.suggestedmodel"]<-predict_Life.expecentancy_suggested
```

```{r}
plot(test$predicted.by.suggestedmodel,test$Life.expectancy)
abline(lm(test$predicted.by.suggestedmodel~test$Life.expectancy))
```

The adjusted r-square of the finalmodel is more than the adjusted r-square of the suggested model.

```{r}
#calculating rmse metric
#install.packages("metrics")
library(ModelMetrics)
rmse_final <- rmse(test$Life.expectancy, test$predicted.by.finalmodel)
rmse_suggested<-rmse(test$Life.expectancy,test$predicted.by.suggestedmodel)
rmse_final
rmse_suggested
```

The RMSE of the full model is less than the RMSE of the suggested model.

By considering both adjusted r square and RMSE values,finalmodel is the best fit for the dataset.

The equation for the multi-linear regression model is-

$\hat{Life.expectancy}=\hat{\beta_0}+\hat{\beta_1}\hat{Alcohol}+\hat{\beta_2}\hat{Total.expenditure}+\hat{\beta_3}\hat{Income.composition.of.resources}+\hat{\beta_4}\hat{Schooling}+\hat{\beta_5}\hat{log(percentage.expenditure)}+\hat{\beta_6}\hat{HIV.AIDS}+\hat{\beta_7}\hat{Hepatitis.B}$
